```{r, warning=FALSE, message=FALSE, error=FALSE, echo = FALSE}
# loading packages into R
library(tidyverse) # data management and visulaization
library(ggpubr) # Supplemental package for ggplot2 that makes pretty tables
```

# (PART) Introduction to Dplyr {-}
# What is dplyr?
<div class = semiemph>
- dplyr is a package that tries to provide a set of functions that utilizes a consistent design, philosophy, grammar, and data structure
- This consistency increases usability and interpretability of code
- It is consistently updated and supported by members of the R-Core team and creaters of RStudio
- It is the most commonly used to manipulate data within the R program
</div>
## Why is Data Manipulation Important?

### Example 1: Survey Data

```{r, echo = FALSE, fig.height=2, fig.width=11}
survey_dat<-read.csv("suppl/survey_data.csv")

survey_dat%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```

### What's Wrong With the Survey Data?
- Some of the meta-data collected by the survey platform is not meaningful.
- It is unclear what the data (i.e., Q1.1) is referring to.
- Items that start with Q1 and Q2 are associated with unique scales that need to be formed into composites.
- Some observations were created by you during pilot testing and should not be included.

## This Isn't Relevant to Me - My Research is Experimental!

### Example 2: Experimental Data
```{r,echo=FALSE, fig.width=7.5, fig.height=2}

set.seed(1234)
ex_data<-tibble(id = 1:6,
                 gender = sample(c("M", "F"), size = 6, replace = TRUE),
                condition = rep(0:1, each = 3),
                pre = round(rnorm(6, 5, sd = 1), 2), 
                non_naieve = c(rep(0, 4), rep(1, 2)))
set.seed(1234)
ex_datb<-tibble(id = c(1, 3:6),
                 gender = sample(c("M", "F"), size = 6, replace = TRUE)[-2],
                condition = c(rep(0, each = 2), rep(1, each = 3)),
                post = round(rnorm(5, 6, sd = 1), 2), 
                non_naieve = c(rep(0, 3), rep(1, 2)))

t1<-ggpubr::ggtexttable(ex_data, theme = ttheme(base_style="lBlueWhite"))
t2<-ggpubr::ggtexttable(ex_datb, theme = ttheme(base_style="lBlueWhite"))


ggarrange(t1,t2)
```

### What's Wrong With the Experimental Data?
- Some of your participants figured out the purpose of your expiriment making their responses invalid.
- Your pre and post scale was miscalibrated and is .3 higher than it should be.
- Your pre and post measures are stored in seperate data files.
- Making matters more difficult, you have 17% participant attrition so you can't just copy and paste data frames together.

## Take-Aways
<div class = activity>
<div class = activitytitle>
Why Does Data Mangaement Matter?</div>
1. Data is messy, no matter what paradigm you work in.
2. Models have different structuring requirements.
3. Knowing how to use a robust set of tools for data management will save you time.
</div>


