```{r, warning=FALSE, message=FALSE, error=FALSE, echo = FALSE}
library(tidyverse)
```

# What is dplyr?
<div class = semiemph>
- dplyr is a package that tries to provide a set of functions that utilizes a consistent design, philosophy, grammar, and data structure
- This consistency increases usability and interpretability of code
- It is consistently updated and supported by members of the R-Core team and creaters of RStudio
- It is the most commonly used to manipulate data within the R program
</div>
## Why is Data Manipulation Important?

### Example 1: Survey Data
<div class = tiny>
```{r, echo = FALSE}
ex_dat<-read.csv("suppl/cons_perf.csv")

knitr::kable(head(ex_dat)%>%select(ResponseId, Status, last_name = RecipientLastName, starts_with("Q")), format = "html", caption = "Example of Survey Data") %>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```
</div>

### What's Wrong With the Survey Data?
- Some of the meta-data collected by the survey platform is not meaningful.
- It is unclear what the data (i.e., Q1.1) is referring to.
- Items that start with Q1 and Q2 are associated with unique scales that need to be formed into composites.
- Some observations were created by you during pilot testing and should not be included.

## This Isn't Relevant to Me - My Research is Experimental!

### Example 2: Experimental Data
```{r,echo=FALSE}
ex_dat2a<-tibble(id = 1:6,
                condition = rep(0:1, each = 3),
                pre = rnorm(6, 5, sd = 1), 
                non_naieve = c(rep(0, 4), rep(1, 2)))

ex_dat2b<-tibble(id = 1:6,
                condition = rep(0:1, each = 3),
                post = rnorm(6, 6, sd = 1), 
                non_naieve = c(rep(0, 4), rep(1, 2)))

knitr::kable(list(head(ex_dat2a), head(ex_dat2b)),digits = 2, format = "html", caption = "Example of Experimental Data" )%>%
  kableExtra::kable_styling()
```

### What's Wrong With the Experimental Data?
- Some of your participants figured out the purpose of your expiriment making their responses invalid.
- Your pre and post scale was miscalibrated and is .3 higher than it should be.
- Your pre and post measures are stored in seperate data files.

## Take-Aways
<div class = semiemph>
1. No matter what paradigm you work in data management is critical.
2. Tidy data rarely exists in the wild.
3. Efficiently managing data will save you time and make you a valuable assest to your collaborators.
</div>


