# Core dplyr Functions

<div class = activity>
<div class = activitytitle>
Core dplyr Functions for Data Manipulation</div>
- filter(): select rows based on some logical condition
- select(): select columns based on their names
- rename(): rename columns
- mutate(): add new variables that are functions of old variables
- group_by(): perform grouped operations
- summarise(): create summary statistics for a group
- arrange(): reorder rows based on some column
</div>

## dplyr Function Structure {#form}

All of the core dplyr functions take the following form:

<div class = activity>
<div class = structure>
function(data, transformation, ...)</div>


1. function: the dplyr function that you want to use

2. data: the data frame or tibble you want to use the function on

3. transformation: the transformation that you want to perform

4. ...: other transformations you want to perform
</div>


## filter(): Retaining Rows {#filter}
- This function allows you to subset the data frame based on a logical test.
- Simply put, it allows you to choose which rows to keep.

### filter() Structure

<div class = activity>
<div class = structure>

filter(data, logical_test, ...)</div>


- Remember, all dplyr functions take the same general form (See section \@ref(form)).
- The first argument specifies the data frame that we are manipulating.
- The second argument specifies the transformation we want to preform.
- In this case transformation argument uses a logical test to define the observations we would like to keep.
- Logical tests can explicitly use logical operators (i.e., == or %in%).
- Functions that return logical values can also be used (i.e., is.na()).
- Multiple logical tests can be provided as indicated by the ellipse.
- If tests are separated by a comma or ampersand, both tests must be TRUE for the observation to be retained.
- If tests are separated by a pipe (i.e., |), either argument can be satisfied for the observation to be retained

</div>

### Using filter()
- Remember the survey data?
- Some observations were created when the survey was being tested. 
- These observations are not informative and should be removed. 
- Luckily, the survey platform records whether a response is from a participant or a tester in the Status column (0 = participant, 8 = tester).
- Using filter(), we can easily retain the real observations while excluding rows associated with the pilot test.

```{r, include = FALSE}
# Creates the survey data we will be working with
# Run this to load the data into the environment
survey_data<-read_csv("suppl/survey_data.csv")
```

```{r, echo=FALSE, fig.height=2, fig.width=8, fig.cap="Raw Data"}
survey_data%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```


```{example, filter1}
Using filter to retain non-pilot observations (Status = 0).
```{r, eval=FALSE}
filter(survey_data, Status == 0)
```

```{r, echo = FALSE, fig.height=2, fig.width=8, fig.cap="Filtered Data"}
## Prints the data
filter(survey_data, Status == 0)%>%
    ggtexttable(theme = ttheme(base_style = "lBlueWhite"))

# Saves it to the enviornment
survey_data<-filter(survey_data, Status == 0)
```


```{example, filter2}
A less practical example that retains observations that responded to Q1.1 OR Q1.3 with 5
```{r, eval=FALSE}
filter(survey_data, Q1.1 == 5 | Q1.3 == 5 )
```

```{r, echo = FALSE, fig.height=1.5, fig.width=8, fig.cap="Participants Who Responded 5 to questions Q1.1 OR Q1.3"}
# Prints the less practical example but doesn't save it
filter(survey_data, Q1.1 == 5 | Q1.3 == 5)%>%
      ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

## select(): Choosing Columns
- Often when cleaning data, we only want to work with a subset of columns. 
- select() is used to retain or remove specific columns. 


### select() Structure

<div class = activity>
<div class = structure>

select(data, cols_to_keep, ...)</div>
- Again, select() takes the general dplyr form (See section \@ref(form)).
- The first argument specifies the data frame that we are manipulating.
- The second argument specifies the transformation we want to preform.
- In this case, the transformation argument specifies a column or columns we would like to keep, separated by commas.
- If you want to keep a range of columns you can specify the first column and last column of the range with a colon.
- Sometimes, it is more efficient to drop then select columns.
- To remove columns, simply include a minus sign in front of the column name.
- select() can also be used to reorder columns - the columns will be ordered how you type them.

</div>

### Useful Helper Functions for select()
- starts_with() used in tandem select() allows you to keep variables that share a stem.
- ends_with() used in tandem with select() allows you to keep variables that share a suffix.
- contains() used in tandem with select() allows you to keep variables that share some common string anywhere in their structure.
- These can be used along with [regular expressions](https://cran.r-project.org/web/packages/stringr/vignettes/regular-expressions.html) to automate large portions of data cleaning.
- Helper functions can speed up the data cleaning process while keeping your code easy to interpret.

### Using select()
- Again, this function helps us solve two issues in the survey data example.
- The survey platform created a column of data for the participant's last name that is completely empty.
- Furthermore, the Status column is no longer informative because all the values should equal 0.
- We can remove this column entirely using the select function.
- All of the following examples complete the same task using different methods although some are more efficient than others!

```{r, echo=FALSE, fig.height=2, fig.width=8, fig.cap="Most Recent Data"}
survey_data%>%
      ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

```{example, select1}
Using select() by specifying columns to retain.

```{r, eval=FALSE}
select(survey_data, ResponseId, Q1.1, Q1.2, Q1.3, Q2.1, Q2.2, Q2.3)
```

```{r, echo = FALSE, fig.height=2, fig.width=8, fig.cap="Selected Data"}
select(survey_data, ResponseId, Q1.1, Q1.2, Q1.3, Q2.1, Q2.2, Q2.3)%>%
      ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

```{example, select2}
Using select() by specifying columns to omit.

```{r, eval=FALSE}
select(survey_data, -Status, -last_name)
```

```{r, echo = FALSE, fig.height=2, fig.width=8, fig.cap="Dropped Data"}
select(survey_data, -Status, -last_name)%>%
      ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

```{example, select3}
Using select() by specifying range of columns.

```{r, eval=FALSE}
select(survey_data, ResponseId, Q1.1:Q2.3)
```

```{r, echo = FALSE, fig.height=2, fig.width=8, fig.cap="Range of Variables Selected"}
select(survey_data, ResponseId, Q1.1:Q2.3)%>%
     ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

```{example, select4}
Using select() with helper functions.

```{r, eval=FALSE}
select(survey_data, contains("id", ignore.case = TRUE), starts_with("Q"))
```

```{r, echo = FALSE, fig.height=2, fig.width=8, fig.cap="Select with Helper Functions"}
select(survey_data, contains("id", ignore.case = TRUE), starts_with("Q"))%>%
    ggtexttable(theme = ttheme(base_style = "lBlueWhite"))

  

survey_data<-select(survey_data, contains("id", ignore.case = TRUE), starts_with("Q"))
```

## rename(): Renaming Variables
- This function is very self explanatory - it renames columns (variables)

### rename() Structure
<div class = activity>
<div class = structure>

rename(data, old_name = new_name, ...)</div>
- Following the general dplyr form (See section \@ref(form)), the first argument specifies the data you are manipulating.
- In this case the transformation arguments take the form of an equation, where the new column name is on the left of the equals sign and the old column name is on the right.
- Multiple variables can be renamed within one rename call, as indicated by the ellipse.
</div>

### Using rename()
- Given that Q1.x and Q2.x are not meaningful stems, we should rename the items so that they are interpretable.
- It turns out that items that are labeled with the prefix "Q1" measured conscientiousness and items that are measured Q2 measure job performance. 

```{r, echo = FALSE, fig.height=2, fig.width=8, fig.cap="Most Recent Data"}
survey_data%>%
    ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```


```{example, rename1}
Using rename() to provide substantive column names.

```{r, eval=FALSE}
rename(survey_data, cons1 = Q1.1, cons2 = Q1.2, cons3 = Q1.3, perf1 = Q2.1, perf2 = Q2.2, perf3 = Q2.3)
```

```{r, echo = FALSE, fig.height=2, fig.width=8, fig.cap="Renamed Data"}
rename(survey_data, cons1 = Q1.1, cons2 = Q1.2, cons3 = Q1.3, perf1 = Q2.1, perf2 = Q2.2, perf3 = Q2.3)%>%
      ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

```{example, rename2, fig.cap = "Renamed Data using Select"}
select() can be used to rename columns as well!

```{r, eval=FALSE}
select(survey_data, ResponseId, cons1 = Q1.1, cons2 = Q1.2, cons3 = Q1.3, perf1 = Q2.1, perf2 = Q2.2, perf3 = Q2.3)
```

```{r, echo = FALSE, fig.height=2, fig.width=8}
select(survey_data, ResponseId, cons1 = Q1.1, cons2 = Q1.2, cons3 = Q1.3, perf1 = Q2.1, perf2 = Q2.2, perf3 = Q2.3)%>%
    ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

```{example, rename3}
rename() may be one area where dplyr is lacking in efficiency. Here is the base R code to do the same task!

```{r, eval=FALSE}
colnames(survey_data)<-c("ResponseId", paste0("cons", 1:3), paste0("perf", 1:3))
survey_data
```

```{r, echo = FALSE, fig.height=2, fig.width=8, fig.cap = "Renamed Data Using Base R"}
survey_data2<-survey_data
colnames(survey_data2)<-c("ResponseId", paste0("cons", 1:3), paste0("perf", 1:3))

survey_data2%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))

survey_data<-rename(survey_data, cons1 = Q1.1, cons2 = Q1.2, cons3 = Q1.3, perf1 = Q2.1, perf2 = Q2.2, perf3 = Q2.3)
```

## mutate(): Creating New Variables
- mutate() creates new variables that are defined by some function or operation.

### mutate() Structure
<div class = activity>
<div class = structure>

mutate(data, new_var = function, ...)</div>
- Again, following the general dplyr form (See section \@ref(form)), the first argument specifies the data you are manipulating.
- The next argument specifies the transformation, which in mutate() defines a new variable.
- To do this, you specify a formula that specifies the name of a new variable on the left of the equals sign and a function that creates the new variable on the right.
- In this notation, function refers to any function or operator that creates a vector of output that is as long as the data frame *or* has a single value.
- Multiple new variables can be created within one mutate() call, but should be separated by commas.
</div>

### Helper Functions
- rowwise(): Applies functions across columns within rows.
- ungroup(): Undoes grouping functions such as rowwise() and group_by() (group_by() will be discussed in Section \@ref(group))

### Using mutate()
- Given that there are two sub scales (i.e., conscientiousness and performance) within our survey data, we can create scale scores for these sets of items. 
- Typically, this is done by averaging the item level data.
- mutate() provides an easy way to do this!

```{r, echo = FALSE, fig.height=2, fig.width=8, fig.cap = "Most Recent Data"}
survey_data%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

```{example, mutate1}
Using mutate() and arithmetic operators to create scale scores with missing data.

```{r, eval=FALSE}
mutate(survey_data, cons = (cons1+cons2+cons3)/3, perf = (perf1+perf2+perf3)/3)
```


```{r, echo = FALSE, fig.height=2, fig.width=8, fig.cap="Arithmatic Scale Scores"}
mutate(survey_data, cons = (cons1+cons2+cons3)/3, perf = (perf1+perf2+perf3)/3)%>%
  mutate_if(is.numeric, round, digits = 2)%>%
    ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```



```{example, mutate2}
Using mutate() and rowwise() to create scale scores while handling missing data (use with caution).

```{r, eval=FALSE}
ungroup(mutate(rowwise(survey_data), cons = mean(c(cons1,cons2,cons3), na.rm = TRUE), perf = mean(c(perf1,perf2,perf3), na.rm = TRUE)))
```


```{r, echo = FALSE, fig.height=2, fig.width=8, fig.cap="Rowwise Scale Scores"}
mutate(rowwise(survey_data), cons = mean(c(cons1,cons2,cons3), na.rm = TRUE), perf = mean(c(perf1,perf2,perf3), na.rm = TRUE))%>%
  mutate_if(is.numeric, round, digits = 2)%>%
    ggtexttable(theme = ttheme(base_style = "lBlueWhite"))

survey_data<-mutate(rowwise(survey_data), cons = mean(c(cons1,cons2,cons3), na.rm = TRUE), perf = mean(c(perf1,perf2,perf3), na.rm = TRUE))%>%
  select(matches("[[:alpha:]]$"))%>%
  ungroup()
```

<div class = list>
- Two new functions are used in the code below. 
- percent_rank() calculates the percentage of observations *less than* an observation.
- cume_dist() calculates the percentage of obseravtions *less than or equal to* an observation.
</div>

```{example, mutate3}
While the above examples illustrate composites, you can also create normalize variables (i.e., percents).

```{r, eval=FALSE}
mutate(survey_data, perf_p = percent_rank(perf), perf_cdf = cume_dist(perf))
```

```{r, echo = FALSE, fig.height=2, fig.width=9, fig.cap="Adding Rank Variables"}
survey_data<-mutate(survey_data, perf_p = percent_rank(perf), perf_cdf = cume_dist(perf))%>%
  mutate_if(is.numeric, round, digits = 2)

survey_data%>%
    ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

- This information can be used to provide useful summarise of distributions.
- I demonstrate how this can be visualized below. 

```{r, fig.width = 10}
p1<-survey_data%>%
  select(perf, perf_p, perf_cdf)%>%
  distinct()%>%
  ggplot(aes(x = perf, y = perf_p))+
  geom_density(stat = "identity", fill = "blue", color = "white")+
  scale_x_continuous(name = "Performance", breaks = 1:5, labels = 1:5, limits = c(1, 5))+
  scale_y_continuous(name = "p(Performance< x)", breaks = c(0, .25, .5, .75, 1), labels = paste0(c(0, .25, .5, .75, 1)*100, "%"))+
  labs(title = "Plot of percent_ranks() Output")
  
p2<-survey_data%>%
  select(perf, perf_p, perf_cdf)%>%
  distinct()%>%
  ggplot(aes(x = perf, y = perf_cdf))+
  geom_density(stat = "identity", fill = "blue", color = "white")+
  scale_x_continuous(name = "Performance", breaks = 1:5, labels = 1:5, limits = c(1, 5))+
  scale_y_continuous(name = "p(Performance<= x)", breaks = c(0, .25, .5, .75, 1), labels = paste0(c(0, .25, .5, .75, 1)*100, "%"))+
  labs(title = "Plot of cume_dist() Output")

ggarrange(p1, p2)
```

- Note that, the cume_dist() can be visulized from raw data (without feature engineering) by using the stat_ecdf
- ecdf stands for empirical cumulative density function

```{r}
survey_data%>%
  ggplot(aes(x = perf))+
  geom_density(stat = "ecdf", fill = "blue", color = "white")+
  scale_x_continuous(name = "Performance", breaks = 1:5, labels = 1:5, limits = c(1, 5))+
  scale_y_continuous(name = "p(Performance<= x)", breaks = c(0, .25, .5, .75, 1), labels = paste0(c(0, .25, .5, .75, 1)*100, "%"))+
  labs(title = "Plot of cume_dist() Output")
```



```{example, mutate4}
You can also create binary indicator variables using conditional logic (i.e. if_else() statements). These indicator variables are sometimes referred to a dummy coded variables or one hot encoding.
```{r, eval=FALSE}
mutate(survey_data, lt_50p = if_else(perf_p<=.5, 1, 0))
```

<div class = activity>
<div class = structure>

if_else(logical_test, value_if_TRUE, value_if_FALSE)</div>
- Used to conditionally operate on dataframe
- Uses two different values or algorithms, depending on a logical test
</div>

```{r, echo = FALSE, fig.height=2, fig.width=9.5, fig.cap = "Adding Dummy Variables"}
mutate(survey_data, lt_50p = if_else(perf_p<=.5, 1, 0))%>%
  mutate_if(is.numeric, round, digits = 2)%>%
    ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

## The Pipe Operator (%>%)
- Notice that while cleaning the survey data we have been typing the data argument multiple times.
- Furthermore, using rowwise(), ungroup(), and mutate() all together makes our code difficult to read!
- Wouldn't it be nice if there was some shorthand way to link functions together?
- Lucky for us there is, and it is call the pipe operator. 
- The pipe operator carries forward the output of the previous function and uses it in the function that follows.
- This allows us to string together multiple functions without retyping the data argument. 

### Structure of the Pipe Operator

<div class = activity>
<div class = structure>

function1(data, transformation, ...)%>%
  function2(transformation)</div>
- The pipe operator carriers forward the output of the previous call and uses it in the subsequent function
- Thus, following any call with %>% will carry forward the output into the subsequent function
- The pipe can be used with base R by using a period as a place holder for the data frame.
</div>

### Using the pipe operator
- Let's use the pipe operator to make our code more readable

```{r, include=FALSE}
survey_data<-read_csv("suppl/survey_data.csv")
```


```{example, pipe1}
Using pipe operator (%>%) to redo what we have done thus far.

```{r, eval=FALSE}
survey_data%>%
  filter(Status == 0)%>%
  select(-Status, -last_name)%>%
  rename(cons1 = Q1.1, cons2 = Q1.2, cons3 = Q1.3, perf1 = Q2.1, perf2 = Q2.2, perf3 = Q2.3)%>%
  rowwise()%>%
  mutate(cons = mean(c(cons1,cons2,cons3), na.rm = TRUE), perf = mean(c(perf1,perf2,perf3), na.rm = TRUE))%>%
  ungroup()%>%
  mutate(perf_p = percent_rank(cons), perf_cdf = cume_dist(perf), lt_50p = if_else(perf_p<=.5, 1, 0))%>%
  select(matches("[[:alpha:]]$"))
```

```{r, echo = FALSE, fig.height=2, fig.width=9.5, fig.cap = "Replicating Data cleaning with Piping"}
survey_data<-survey_data%>%
  filter(Status == 0)%>%
  select(-Status, -last_name)%>%
  rename(cons1 = Q1.1, cons2 = Q1.2, cons3 = Q1.3, perf1 = Q2.1, perf2 = Q2.2, perf3 = Q2.3)%>%
  rowwise()%>%
  mutate(cons = mean(c(cons1,cons2,cons3), na.rm = TRUE), perf = mean(c(perf1,perf2,perf3), na.rm = TRUE))%>%
  ungroup()%>%
  mutate(perf_p = percent_rank(perf), perf_cdf = cume_dist(perf), lt_50p = if_else(perf_p<=.5, 1, 0))%>%
  select(matches("[[:alpha:]]$"))

survey_data%>%
  mutate_if(.predicate = is.numeric, round, digits = 2)%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

## group_by(): Grouping Data Frames {#group}

```{r, echo = FALSE}
manager<-read_csv("suppl/Manager Data.csv")

survey_data<-left_join(survey_data, manager)%>%
  select(ResponseId, Manager, everything())
```

- Sometimes, when working with data we want to perform some operation within a grouping variable.
- For example, the participants responding to this survey report to different managers.
- We may be interested in creating a new column of data that contains the work-groups' average performance.
- group_by() can be used in tandem with mutate() to apply a function within columns clustering on groups

### group_by Structure

<div class = activity>
<div class = structure>

group_by(data, grouping_variable, ...)</div>
- group_by() takes the common dplyr structure - define the data and then define the transformation.
- The transformation in this case simply defines the grouping variable.
- If multiple grouping variables are provided, the data is grouped by unique combinations of all grouping variables.
- Note that this function is similar to rowwise() in that no physical change happens to the data - it only affects how later functions act the object. 
- Because of this, group_by() is rarely (dare I say never) used without being accompanied by other functions such as mutate() or summarise() (to be covered in Section \@ref(sum))
- Also, just like rowwise(), in order return the data set to its ungrouped form it is necessary to call the ungroup() function after finishing grouped manipulations.

</div>

### Using group_by()
- The survey data has been joined with information regarding employees managers.
- We can now calculate each employee's team's average performance, conscientiousness, and the number of teammates who responded in the data.
- While I only illustrate how to use group_by() with the pipe operator, if for some reason you wanted to use a single group_by() call instead of a chain, it can be done.

```{r, echo = FALSE, fig.height=2, fig.width=10, fig.cap = "Cleaned Data with Manager Info"}
survey_data%>%
  mutate_if(is.numeric, round, digits = 2)%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

```{example, groupby1}
Using group_by() to create team level variables and n() to create group size variables.
```{r, eval = FALSE}
survey_data%>%
  group_by(Manager)%>%
  mutate(team_cons = mean(cons, na.rm = TRUE), team_size = n())%>%
  ungroup()
```

```{r, echo = FALSE, fig.height=2, fig.width=12, fig.cap="Group and Size Variables"}
survey_data<-survey_data%>%
  group_by(Manager)%>%
  mutate(team_cons = mean(cons, na.rm = TRUE), team_size = n())%>%
  mutate_if(is.numeric, round, digits = 2)%>%
  ungroup()

survey_data%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```
```{example, groupby2}
Using group_by() to create team level cumulative distribution.
```{r, eval = FALSE}
survey_data%>%
  group_by(Manager)%>%
  mutate(team_cdf = cume_dist(perf))%>%
  ungroup()
```

```{r, echo = FALSE, fig.height=2, fig.width=11.5, fig.cap="Within-team CDF"}
tmp_survey_data<-survey_data%>%
  group_by(Manager)%>%
  mutate(team_cdf = cume_dist(perf))%>%
  ungroup()%>%
  mutate_if(is.numeric, round, digits = 2)

tmp_survey_data%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

The CDF can be used to visulize distributional differences across groups as well. 

```{r, echo = FALSE}
tmp_survey_data%>%
  select(Manager, perf, team_cdf)%>%
  distinct()%>%
  ggplot(aes(x = perf, y= team_cdf, fill = Manager))+
  geom_density(stat = "identity")+
  facet_wrap(~Manager)+
  scale_x_continuous(name = "Performance", breaks = 1:5, labels = 1:5, limits = c(1, 5))+
  scale_y_continuous(name = "p(Performance<= x)", breaks = c(0, .25, .5, .75, 1), labels = paste0(c(0, .25, .5, .75, 1)*100, "%"))+
  labs(title = "Plot of Grouped cume_dist() Output")
```


```{example, addcount}
add_count() is a nice alternative, to the group_by()%>%mutate() chain if your goal is to simply add a grouped frequency variables to the data frame.
```{r, eval = FALSE}
survey_data%>%
  add_count(Manager)
```

```{r, echo = FALSE, fig.height=2, fig.width=8, fig.cap="Alternative method for creating count variables"}
survey_data%>%
  add_count(Manager)%>%
  mutate_if(is.numeric, round, digits = 2)%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```


## summarise(): Creating Data Summaries {#sum}

- While creating grouped variables is sometimes necessary for analyses, often we simply want to describe properties of our data.
- summarise() is especially useful for this because it applies a function across rows of data to create a single value.
- If the data is grouped, there is a value returned for each group.


### summarise() Structure

<div class = activity>
<div class = structure>

summarise(data, summary_var = function, ...)</div>

- Following the consistent dplyr structure, summarise() requires that you first specify the data and then a transformation.
- The transformation in summarise takes a similar form as mutate().
- The left hand side of the equation defines the name of a new summary variable and the right hand side defines a function or operation.
- The function should return a single value (i.e., mean() or sd()).
</div>

### Using summarise()
- Let's create a summary table for the overall sample as well as each team


```{example, sum1}
Using summarise() to create a summary table for the entire survey data frame
```{r, eval = FALSE}
survey_data%>%
  summarise(mean_cons = mean(cons, na.rm = TRUE), mean_perf = mean(perf, na.rm = TRUE),
            sd_cons = sd(cons, na.rm = TRUE), sd_perf = sd(perf, na.rm = TRUE))
```

```{r, echo = FALSE, fig.height=1, fig.width=12, fig.cap = "Summary Statistics"}
survey_data%>%
  summarise(mean_cons = mean(cons, na.rm = TRUE), mean_perf = mean(perf, na.rm = TRUE),
            sd_cons = sd(cons, na.rm = TRUE), sd_perf = sd(perf, na.rm = TRUE))%>%
  mutate_if(is.numeric, round, digits = 2)%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))

```

```{example, sum2}
Using group_by() and summarise() to create a summary table for different work groups
```{r, eval = FALSE}
survey_data%>%
  group_by(Manager)%>%
  summarise(team_cons = mean(cons, na.rm = TRUE), team_perf = mean(perf, na.rm = TRUE),
            sd_cons = sd(cons, na.rm = TRUE), sd_perf = sd(perf, na.rm = TRUE))
```

```{r, echo = FALSE, fig.height=1, fig.width=12, fig.cap="Grouped Summary Statistics"}
survey_data%>%
  group_by(Manager)%>%
  summarise(team_cons = mean(cons, na.rm = TRUE), team_perf = mean(perf, na.rm = TRUE),
            sd_cons = sd(cons, na.rm = TRUE), sd_perf = sd(perf, na.rm = TRUE))%>%
    mutate_if(is.numeric, round, digits = 2)%>%
   ggtexttable(theme = ttheme(base_style = "lBlueWhite"))

```

```{example, count}
count() is a nice alternative to the group_by()%>%summarise() chain if your goal is simply to describe grouped frequencies.
```{r, eval = FALSE}
survey_data%>%
  count(Manager)
```

```{r, echo = FALSE, fig.height=1, fig.width=12, "Grouped Frequencies"}
survey_data%>%
  count(Manager)%>%
    mutate_if(is.numeric, round, digits = 2)%>%
   ggtexttable(theme = ttheme(base_style = "lBlueWhite"))

```

## arrange(): Ordering Rows
- arrange() can be used to sort rows in a data frame
- By default, arrange() orders a data frame from values in a column that go from smallest to largest
- You can use desc() with arrange to sort from largest to smallest

### arrange() Structure
<div class = activity>
<div class = structure>
arrange(data, sort_var, ...) </div>
- arrange() takes the same structure as all other core dplyr functions.
- First, specify the data you are manipulating
- Second, specify the transformation - the column or columns you are sorting by
- If multiple columns are provided, arrange will sort by the first column and use subsequent columns as tie breakers
</div>


### Using arrange()

```{example, arr1}
Adding arrange() to our summary table
```{r, eval = FALSE}
survey_data%>%
  group_by(Manager)%>%
  summarise(team_cons = mean(cons, na.rm = TRUE), team_perf = mean(perf, na.rm = TRUE),
            sd_cons = sd(cons, na.rm = TRUE), sd_perf = sd(perf, na.rm = TRUE))%>%
  arrange(team_cons)
```

```{r, echo = FALSE, fig.height=1, fig.width=8, fig.cap="Sorting the Summary Table"}
survey_data%>%
  group_by(Manager)%>%
  summarise(team_cons = mean(cons, na.rm = TRUE), team_perf = mean(perf, na.rm = TRUE),
            sd_cons = sd(cons, na.rm = TRUE), sd_perf = sd(perf, na.rm = TRUE))%>%
  arrange(team_cons)%>%
  mutate_if(is.numeric, round, digits = 2)%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))

```

```{example, arr2}
Sorting the summary table in descending order with desc() and arrange()
```{r, eval = FALSE}
survey_data%>%
  group_by(Manager)%>%
  summarise(team_cons = mean(cons, na.rm = TRUE), team_perf = mean(perf, na.rm = TRUE),
            sd_cons = sd(cons, na.rm = TRUE), sd_perf = sd(perf, na.rm = TRUE))%>%
  arrange(desc(team_cons))
```

```{r, echo = FALSE, fig.height=1, fig.width=8, fig.cap="Sorting in Descending Order"}
survey_data%>%
  group_by(Manager)%>%
  summarise(team_cons = mean(cons, na.rm = TRUE), team_perf = mean(perf, na.rm = TRUE),
            sd_cons = sd(cons, na.rm = TRUE), sd_perf = sd(perf, na.rm = TRUE))%>%
  arrange(desc(team_cons))%>%
  mutate_if(is.numeric, round, digits = 2)%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

