# Join Functions

- Let's look at the original survey data set after it was cleaned up. 

```{r, echo = FALSE}
ex_data<-head(ex_dat)%>%select(ResponseId, Status, last_name = RecipientLastName, starts_with("Q"))
ex_data%>%
  filter(Status == 0)%>%
  select(-Status, -last_name)%>%
  rename(cons1 = Q1.1, cons2 = Q1.2, cons3 = Q1.3, perf1 = Q2.1, perf2 = Q2.2, perf3 = Q2.3)%>%
  rowwise()%>%
  mutate(cons = mean(c(cons1,cons2,cons3), na.rm = TRUE), perf = mean(c(perf1,perf2,perf3), na.rm = TRUE))%>%
  knitr::kable(digits = 2)%>%
  kableExtra::kable_styling(bootstrap_options = "striped", full_width = FALSE, position = "left")
```

- Note that manager information was not originally stored in this data - a data set containing manager information had to be merged with the original data frame.

```{r, echo = FALSE}
man_data<-rbind(head(ex_dat)%>%
  mutate(Manager = c("David","Nick", "Julia", "Nick", "Nick", "Julia"))%>%
  select(ResponseId, Manager), 
  head(ex_dat)%>%
  mutate(Manager = c("David","David", "Julia", "Nick", "Nick", "Julia"), 
         ResponseId=str_replace(ResponseId, "[[2:7]]", "4"))%>%
  select(ResponseId, Manager))

  knitr::kable(man_data)
```


- Merging data sets can be rather complicated, as you will soon see. 
- Luckily the creators of dplyr wrote a set of functions that make merging multiple data tables easy.

## Overview of dplyr Functions for Joining and Binding

### Binding Functions
- bind_rows(): Stacks many data frames vertically. 
- bind_cols(): Joins many date frames horizontally.

### Mutating Joins
- left_join(): Joins based on an ID variable. Retains all rows in left data frame and only matching rows in the right.
- right_join(): Joins based on an ID variable. Retains all rows in the right data frame and only matching rows in the left.
- inner_join(): Joins based on an ID variable. Retains only matching rows for both data frames.
- full_join(): Joins based on an ID variable (Considers order). Retains only matching rows for 

### Filtering Joins
- semi_join(): Retains rows in the left hand data frame that match an ID variable in the right hand data frame. 
- anit_join(): Retains rows in the left hand data frame that do NOT match and ID variable in the right hand data frame. 


## bind_cols(): Binding Data Frames Horizontally
- bind_cols() is used when you have a set of data frames that 

    1. Have equal number of rows
    2. Are ordered identically, with no missing or new observations
    
- If these two requirements are not met, the data will joined combining information about different participants
- If your data does not meet either of these requirements, but has a participant identifier use a join function discussed below.
    
### bind_cols() Structure

<div class = activity>
<div class = activitytitle>

bind_cols(...)

</div>
- bind_cols() takes two or more data frames (or a list of data frames) that have an equal number of identically ordered rows.
</div>

### Using bind_cols()
- Note that, while for the survey data we want to combine two data frames horizontally, the data frames do not have the same number of rows.
- Furthemore, based on the responseId variable, we know that participants are not in the same order.
- Thus we are probably better off using a different function to join these two data sets. 
- In contrast, note that in the two tabels below, each data frame has the same number of observations and the ID variables align perfectly.

```{r, echo = FALSE}
cb1_data <- ex_data %>%
  filter(Status == 0) %>%
  rename(
    cons1 = Q1.1,
    cons2 = Q1.2,
    cons3 = Q1.3,
    perf1 = Q2.1,
    perf2 = Q2.2,
    perf3 = Q2.3
  ) %>%
  rowwise() %>%
  mutate(cons = mean(c(cons1, cons2, cons3), na.rm = TRUE),
         perf = mean(c(perf1, perf2, perf3), na.rm = TRUE))%>%
               select(ResponseId, cons)

cb2_data <- ex_data %>%
  filter(Status == 0) %>%
  rename(
    cons1 = Q1.1,
    cons2 = Q1.2,
    cons3 = Q1.3,
    perf1 = Q2.1,
    perf2 = Q2.2,
    perf3 = Q2.3
  ) %>%
  rowwise() %>%
  mutate(cons = mean(c(cons1, cons2, cons3), na.rm = TRUE),
         perf = mean(c(perf1, perf2, perf3), na.rm = TRUE))%>%
               select(ResponseId, perf)

knitr::kable(list(cb1_data,
             cb2_data), digits = 2, caption = "Two Tables With Corresponding Rows and Equal Dimensions")
```

```{example, bindc}
Binding data frames together horizontally.
```{r, eval = FALSE}
bind_cols(cons_dat, perf_dat)
```

```{r, echo = FALSE}
bind_cols(cb1_data, cb2_data)%>%
  knitr::kable(digits = 2, caption = "Output of bind_cols()")
```

- Note that, since there is a dupicate column (ResponseId) in the new joined dataset, col_bind() automatically added a 1 to the end of the column name.
- This is intended to prevent mixups, but can result in dupicate data.
- While this example was adequate for illustration purposes, in practice, join functions are more flexible and appropriate when data sets have  shared identifier.

## bind_rows(): Binding Data Frames Vertically
- bind_rows() is used when you want to bind data frames vertically.
- This is sometimes referred to as stacking data frames.
- Unlike bind_cols(), bind_rows() attempts to match columns based on their names.
- If a data frame is missing a column, observations will have missing data for that variable.

### bind_rows() Structure

<div class = activity>
<div class = activitytitle>

bind_rows(...)

</div>

- bind_rows() takes two or more data frames (or a list of data frames) 

</div>

### Using bind_rows()

```{r, echo = FALSE, message=FALSE, warning=FALSE}
br_data<-ex_data%>%
          filter(Status == 0) %>%
          rename(
            cons1 = Q1.1,
            cons2 = Q1.2,
            cons3 = Q1.3,
            perf1 = Q2.1,
            perf2 = Q2.2,
            perf3 = Q2.3
          ) %>%
          rowwise() %>%
          mutate(cons = mean(c(cons1, cons2, cons3), na.rm = TRUE),
                 perf = mean(c(perf1, perf2, perf3), na.rm = TRUE))%>%
          select(ResponseId, cons, perf)
          
br1_data<-br_data[1:3,]
br2_data<-br_data[4:nrow(br_data),]

knitr::kable(list(br1_data, 
                  left_join(br2_data,man_data)%>%
  select(Manager, ResponseId, perf, cons)), digits = 2, caption="Two data frames with a different number of columns ordered differently.")

```

```{example, bindr}
Binding data frames together vertically (AKA, stacking).
```{r, eval = FALSE}
bind_rows(cons_dat, perf_dat)
```

```{r, echo = FALSE, message = FALSE, warning=FALSE}
knitr::kable(bind_rows(br1_data, left_join(br2_data,man_data)%>%
                  select(Manager, ResponseId, perf, cons)),
  digits = 2, caption = "Output of bind_cols()")
```

- Note that columns that are named the same are appropriately matched. 
- Furthermore, observations from a data frames that is missing a column are assigned NA for that variable.






