# (PART) Advanced Dplyr {-}

# Other Functions for Extracting Observations

- In the previous chapters, we learned several functions that can be used to extract obseravtions from a data frame. 
- filter() uses a logical test to extract observations.
- In contrast, filtering joins use an ID variable in another data frame to extract observations.
- In the chapters that follow we will cover functions that allow us to extract observations when we want to 

<div class=list>
1. Want to take a random subset of observations
2. Want to retain only distinct observations
</div>


## Random Samples of Observations
- While very much beyond the scope of this lecture, randomly splitting a data set is a key operation for many statistical procedures. 
- Researchers who want to cross-validated their models subset their data (sometimes $n$ times) to evaluate its performance
- When a model's statistical assumptions are violated, a researcher can implement a procedure called bootstrapping that uses resampling methods to estimate a parameters standard error. 
- In short, knowing how to randomly sample a data set opens the door to many other statistical procedures
- The following sections will define the structural form, and show a few examples, however this section is not ended to go indepth into resampling methods. 

### Structural Form of sample_ Functions
<div class = activity>
<div class = structure>
sample_x(data, size, replace, weight, ...)</div>
- sample_x denotes the sampling function you would like to use.
- data: specifies the data frame you would like to operate on.
- size: specifies the size of the sample you would like to take either in absolute or relative terms (depending on whether you use sample_n() or sample_frac()).
- replace: logical value specifying whether a data frame should be sampled with replacement (i.e., bootstrap). 
- weight: a vector of weights equal to the number of observations in the dataframe specifying how likely each observation is to be sampled (useful for stratefied sampling). 
</div>

### Using sample_n() and sample_frac()
- sample_frac() and sample_n() only differ in terms of the size argument
- For sample_frac() you specify a proportion, relative to data argument. 
- For sample_n() you specify an absolute number of rows for the output data. 
- Arguably, sample_frac() is more robust to changes in upstream code. 
- For example, if you catch a mistake in your data cleaning prior to taking a random sample of your data set, sample_frac() will still sample relative to this new data frame. 
- In contrast, sample_n() does not adjust to changes in your code and will still resample based on the n you define.
- Putting differences aside, lets consider how a researcher could create a training and test data frame to evaluate their model's performance.
- When randomly sampling the data frame, always make sure to use set.seed() so that the results are reproducible.
- I will also include and example that shows you how to work around the potential pitfalls fo sample_n() by avoiding hardcoding its size argument.


```{r, echo=FALSE, fig.height=3, fig.width=8}
ex_data%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```


```{example, samplen1}
Using sample_n() and anti_join() to create training and test sets, hardcoding size.
```{r, eval = FALSE}
set.seed(123)
training <- sample_n(ex_data, size = 4, replace = FALSE)
holdout <- anti_join(ex_data, training)
```

```{r, fig.width=14, fig.height=2, message=FALSE, echo=FALSE, warning=FALSE, }
set.seed(123)

training <- sample_n(ex_data, size = 4, replace = FALSE) 
holdout <- anti_join(ex_data, training)

t1<-training%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))


t2<-holdout%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))

ggarrange(t1, t2)
```

```{example, samplefrac}
Using sample_frac() and anti_join() to create training and test sets.
```{r, eval = FALSE}
set.seed(123)
training <- sample_frac(ex_data, size = .8, replace = FALSE)
holdout <- anti_join(ex_data, training)
```

```{r, fig.width=14, fig.height=2, message=FALSE, echo=FALSE, warning=FALSE, }
set.seed(123)

training <- sample_frac(ex_data, size = .8, replace = FALSE) 
holdout <- anti_join(ex_data, training)

t1<-training%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))


t2<-holdout%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))

ggarrange(t1, t2)

```


```{example, samplen2}
Using sample_n() and anti_join() to create training and test sets, without hardcoding n.
```{r, eval = FALSE}
set.seed(123)
rel_n<- nrow(ex_data)*.8
training <- sample_n(ex_data, size = rel_n, replace = FALSE)
holdout <- anti_join(ex_data, training)
```

```{r, fig.width=14, fig.height=2, message=FALSE, echo=FALSE, warning=FALSE}
set.seed(123)

rel_n<-nrow(ex_data)*.8

training <- sample_n(ex_data, size = rel_n, replace = FALSE) 
holdout <- anti_join(ex_data, training)

t1<-training%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))


t2<-holdout%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))

ggarrange(t1, t2)
```


## distinct(): extracting unique observations
- As we know, data collection methods are not perfect and neither are participants. 
- Sometimes, software accidently records duplicate observations. 
- Furthermore, participants may take surveys more than once resulting in duplicate information for the same person.
- Identifying distinct responses is often of critical step in ensuring high fidelity data.
- The final observation extraction function we will cover provides a means of extracting unique cases.

### distinct() Structure
<div class = activity>
<div class = structure>
distinct(data, distinct_var, ..., .keep_all)</div>
- data: specifies the data frame you would like to operate on.
- distinct_var: defines a variable that you would like to identify distinct levels of.
- If multiple variables are provided, distinct identifies unique combinations of these levels.
- .keep_all: is a logical values that specifies whether or not to keep all other variables in the resulting output.
- Note that distinct() returns the first observations with a distinct level of distinct_var.
- This means that if .keep_all = TRUE is only really appropriate when an observation is a true duplicate (all information is redundant).
</div>

### Using distinct()
```{r, fig.height=3.5, fig.width=3, echo = FALSE}
man_data%>%
    ggtexttable(theme = ttheme(base_style="lBlueWhite"))

```

- Let's take a look at the manager data that we used when combining multiple data frames.
- Let's use distinct() create data frames that:

<div class = list>
1. Store the names of each manager.
2. Store the IDs associated with each employee.
3. Store ID/manager combinations.
</div>


```{example, distinct1}
Using distinct() to extract unique levels of manager.
```{r, eval = FALSE}
man_data%>%
  distinct(Manager)
```

```{r, fig.height=1.5, fig.width=3, echo = FALSE}
man_data%>%
  distinct(Manager)%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))

```

```{example, distinct2}
Using distinct() to extract unique levels of ResponseId.
```{r, eval = FALSE}
man_data%>%
  distinct(Employee)
```

```{r, fig.height=3.5, fig.width=3, echo = FALSE}
man_data%>%
  distinct(ResponseId)%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))

```
```{example, distinct3}
Using distinct() to extract unique ResponseId-Manager combinations.
```{r, eval = FALSE}
man_data%>%
  distinct(Employee, Manager)
```

```{r, fig.height=3.5, fig.width=3}
man_data%>%
  distinct(ResponseId, Manager)%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```