# Performing Repeated Operations
- By this point, my hope is that you feel comfortable with the core dplyr functions.
- In the next chapters, we will discuss three variations on each of these functions that allow you to write one line of code to repeat a manipulation many times. 
- While I will not demonstrated every variation of these functions, knowing they exist can help you speed up your data cleaning.
- What are some instances when you would want to do repeated manipulations?
<div class = list>
- Renaming all all columns so that their names are lower case.
- Centering a set of independent variables for regression.
- Calculating summary statistics for a large set of variables.
- Converting a set of variables that were read as character to numeric.
- Rounding all numeric variables to the second decimal place.
</div>

- Most of the core functions have variations that variations that facilitate repeated operations in different ways.
- The names are mostly the same, except a suffix is added to the end to differentiate it from its typical call (i.e., mutate_all())
- Each suffix defines the repeated manipulation in a different way.

<div class = activity>
<div class = activitytitle>
Summary of Suffix Definitions
</div>
- _all: Applies the transformation to all columns.
- _at: Applies the transformation to a set of columns you define.
- _if: Applies the transformation to a set of columns that match an argument.
</div>

## all() suffix
- As you might suspect, functions with the _all() suffix apply your specified transformation to **all** columns in the data frame.
- This is useful when there is a single transformation that is appropriate for every column.
- For example, because R is case sensitive, it is much easier to always were with lower case column names.
- Using rename_all() will help us rename every single column so that it matches this pattern.
- To use rename_all() we simply define the data frame and then a function that will take a column name and change it in some way.
- tolower() converts string values to lowercase and is the appropriate function to use here.

```{r, echo = FALSE, fig.width=10, fig.height=2}
survey_data%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```


```{example, renameall}
Using rename_all() and tolower() to rename all variables so they are lower case.
```{r, eval = FALSE}
survey_data%>%
  rename_all(tolower)
```

```{r, echo=FALSE, fig.height=2, fig.width=8}
survey_data%>%
  rename_all(tolower)%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```

## at() suffix
- The at suffix applies a given transformation to a set of variables.
- It relies on the vars() helper function to define these variables. 
- Let's use the mutate_at() function to center the cons and perf columns
- In this case, I want to retain my centered and uncentered variables for later use.
- To do this, I defined a named list that contains the functions I want to apply to the variables I define.
- The names of the elements in the list will be appended to my original names to create new columns. 
- .s are used as place holders for the vars.

```{r, echo = FALSE, fig.width=10, fig.height=2}
survey_data%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```

```{example, mutateat}
Using mutate_at() to center cons and perf at 0.
```{r, eval = FALSE}
centered_data<-survey_data%>%
  mutate_at(vars(cons, perf), list(c = ~ .-mean(., na.rm = TRUE)))

centered_data
```

```{r, echo=FALSE, echo = FALSE, fig.height=2, fig.width=10}
survey_data%>%
  mutate_at(vars(cons, perf), list(c = ~ .-mean(., na.rm = TRUE)))%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```
- Lets double check our work using summarise_at()
- Centering a variables changes it's expected value but not its spread.
- This means that the new variables should have different means but identical standard deviations compared to the original variables.

```{example, summariseat}
Using summarise_at() to verify the transformation worked appropriately
```{r, eval = FALSE}
centered_data%>%
  summarise_at(vars(cons, perf, cons_c, perf_c), list(mean = ~mean(., na.rm = TRUE),
                                                      sd = ~sd(., na.rm = TRUE)))

```

```{r, echo=FALSE, fig.width=12, fig.height=1}
survey_data%>%
  mutate_at(vars(cons, perf), list(c = ~ .-mean(., na.rm = TRUE)))%>%
  summarise_at(vars(cons, perf, cons_c, perf_c), list(mean = ~mean(., na.rm = TRUE),
                                                      sd = ~sd(., na.rm = TRUE)))%>%
  mutate_if(is.numeric, round, digits = 5)%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```


## if() suffix
- The _if suffix applies a transformation to a set of columns that satisfy a logical test.
- I have been using mutate_if() and round() behind the scenes while writing this book to format most of the tables that you see. 
- Consider Example \@ref(exm:summariseat) when I don't use mutate_if(). 

```{example, summariseat2}
The code you saw really generates this ugly beast. 
```{r, eval = FALSE}
centered_data%>%
  summarise_at(vars(cons, perf, cons_c, perf_c), list(mean = ~mean(., na.rm = TRUE),
                                                      sd = ~sd(., na.rm = TRUE)))

```

```{r, echo=FALSE, fig.width=12, fig.height=1}
survey_data%>%
  mutate_at(vars(cons, perf), list(c = ~ .-mean(., na.rm = TRUE)))%>%
  summarise_at(vars(cons, perf, cons_c, perf_c), list(mean = ~mean(., na.rm = TRUE),
                                                      sd = ~sd(., na.rm = TRUE)))%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```

- The floating point (maximum decimal point) in R goes out a long way making calculations very precise but output unwildy.
- By using mutate_if(), we can apply round() repeatedly across the data frame. 
- round() only works with numeric data, though, so we want to avoid applying it character and factor data.

```{example, summariseat3}
Using summarise_at() to verify the transformation worked appropriately
```{r, eval = FALSE}
centered_data%>%
  summarise_at(vars(cons, perf, cons_c, perf_c), list(mean = ~mean(., na.rm = TRUE),
                                                      sd = ~sd(., na.rm = TRUE)))%>%
  mutate_if(is.numeric, round, digits = 5)
```

```{r, echo=FALSE, fig.width=12, fig.height=1}
survey_data%>%
  mutate_at(vars(cons, perf), list(c = ~ .-mean(., na.rm = TRUE)))%>%
  summarise_at(vars(cons, perf, cons_c, perf_c), list(mean = ~mean(., na.rm = TRUE),
                                                      sd = ~sd(., na.rm = TRUE)))%>%
  mutate_if(is.numeric, round, digits = 5)%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```
