# (PART) Tidy Data with tidyr {-}

# Wider and Longer Data Formats

- You now know how to efficiently add/remove columns, remove rows, and summarise information contained within a data frame. 
- There are a few more tools you need to learn before you can handle most data management tasks. 
- You have seen one form longitudinal data can take in  Chapter 1 (separate objects). 
- It can also come in long format and wide format, depicted below. 

- Transitioning back and forth between these formats is an important skill to have, because different analyses require different data formats. 
- For example, most multi-level modeling software take the data in long format. In contrast, many MANCOVA packages require wide format data. 
- Luckily, tidyr, another package written by Hadley Wickham, is specially designed to reshape data. 
- tidyr contains a set of functions for wrangling messy data and making it tidy. 
- tidy data has the following properties.


    1. Each variable forms a column.
    2. Each observation forms a row.
    3. Each type of observational unit forms a table.

- Please see the [tidyr vignette](https://cran.r-project.org/web/packages/tidyr/vignettes/tidy-data.html) for more information on tidy data.

<div class = activity>
<div class = activitytitle>
Core tidyr Functions for Reshaping Data</div>
- gather(): Make a data frame longer
- spread(): Make a data frame wider
</div>

```{r, fig.cap="Wide Format", fig.height=2, fig.width=4.5, echo = FALSE, message=FALSE}
wide<-ex_data%>%
  left_join(ex_datb)
  
wide%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```

```{r, echo = FALSE, fig.cap="Long Format", message=FALSE}
long<-ex_data%>%
  left_join(ex_datb)%>%
  gather(key = "time", value = "test", pre, post)

long%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```



## gather(): Wider to Longer

gather() is a function intended to take wide format data and make it longer. It does this by gathering a set of columns in the wide data into one column. The user defines the columns columns that should be collapsed, the name of the column that will store the original columns' names, and the name of the column that will store the original columns' values. 

### gather() Structure
<div class = activity>
<div class = structure>

gather(data, key, value, ...)</div>

- key: Name of column to store wide format column names.
- value: Name of column to store the selected columns values.
- ...: A selection of columns to gather into long format. This operates similar to select() and can accommodate special operators like :, -, starts_with, etc.
</div>




### Using gather()
- gather() is useful for converting a data frame into a longer format.
- We will illustrate this with the wide data set shown previously
- This data frame has to columns for a test core taken at two different time points
- These are labeled "pre" and "post"
- gather() can be used to stack the pre and post columns into a single test column which we will define with the value
- The temporal information can be stored in a separate column which we define with the key argument

```{r, echo = FALSE, fig.cap="Wide Format Data"}
wide%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```

```{example, gather1}
We can specify the columns we want to gather using the elipse argument
```{r, eval=FALSE}
gather(wide, key = "time", value = "test", pre, post)
```


```{r, echo = FALSE, fig.cap = "Using gather() to Go From Wide to Long"}
gather(wide, key = "time", value = "test", pre, post)%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```
```{example, gather2, fig.cap = "Alternative column selection"}
Equivalent sytax would be to use - to deselect columns to be gathered. In some use cases, this is quicker.
```{r, eval=FALSE}
gather(wide, key = "time", value = "test", -c(id, gender, condition, non_naieve))
```
## spread(): Longer to Wider

- spread() is gather()'s conjugate. 
- It converts long data frames into wider data frames. 
- To do this, the user specifies a key column, that stores variable names, and a value column that stores the information associated with those variables. 
- The key column is spread so that each unique value stored within it becomes a new column storing the associated values. 

### spread() Structure
<div class = activity>
<div class = structure>

spread(data, key, value)</div>

- key: Name of column to store wide format column names.
- value: Name of column to store the selected columns values.
</div>

### Using spread()

```{r, echo = FALSE, fig.cap="Long Data"}
long%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```

```{example, spread1}
```{r, eval=FALSE}
spread(wide, key = "time", value = "test")
```


```{r, echo = FALSE, fig.cap="Using Spread to Go From Long to Wide"}
spread(long, key = "time", value = "test")%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

## Recent Developments

- The tidyverse is under constant development by a team of very smart people. 
- Recently, Hadley Wickham, one of the key contributes to the tidyverse announced new functions that will come to replace gather() and spread.
- To prepare you for this eventuality, I am going to introduce pivot_wider() and pivot_longer().
- Given that they are still in development, I will only cover the very basics of these functions.
- They are only available in the tidyr development version and are not yet available in the tidyverse package
- To download the development version restart your R session (ctr+shift+f10) and run the following code


```{r, eval = FALSE}
install.packages("devtools")
devtools::install_github("tidyverse/tidyr")
```

### The Problems with gather() and spread()
- Gather and spread will throw errors when working with some data types.
- Some find the functions non-intuitive.
- The latest iteration of these functions make them more robust and also more intuitive.
- Cannot extract meta-data from column names.

## pivot_longer(): gather's() predecessor 
- pivot_longer() is the most recent iteration of gather()
- It takes a data set and transforms it so it is longer, just like gather().

### pivot_longer() structure
<div class = activity>
<div class = structure>

pivot_longer(data, names_to, values_to, cols, ...)</div>
- cols: The columns to gather, defined similar to select() and gather()
- names_to: Column name to store the former column names.
- values_to: Column name to store the former value names.
- ...: to see additional arguments run ?pivot_longer.
</div>


### Using pivot_longer()

```{r, echo = FALSE, fig.cap="Wide Data"}
wide%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```

```{example, pivotlonger}
We can specify the columns we want to gather using the elipse argument
```{r, eval=FALSE}
pivot_longer(wide, cols = c(pre, post), key = "time", value = "test")
```


```{r, echo = FALSE, fig.cap = "Using pivot_longer() to go from Wide to Long"}
pivot_longer(wide, cols = c(pre, post), names_to = "time", values_to = "test")%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```

## pivot_wider(): spread()'s predecessor
- pivot_wider() is the most recent iteration of spread().
- pivot_wider() takes a data frame and makes it wider.

### pivot_wider() Structure
<div class = activity>
<div class = structure>

pivot_wider(data, names_from, values_from, ...)</div>
- names_from: Column name to store the former column names.
- values_from: Column name to store the former value names.
- ...: to see additional arguments run ?pivot_longer.
</div>


```{r, echo = FALSE, fig.cap="Long Data"}
long%>%
  ggtexttable(theme = ttheme(base_style="lBlueWhite"))
```

```{example, spread1}
```{r, eval=FALSE}
pivot_wider(long, names_from =  "time", values_from = "test")
```


```{r, echo = FALSE, fig.cap="Using pivot_wider() to Go From Long to Wide"}
pivot_wider(long, names_from =  "time", values_from = "test")%>%
  ggtexttable(theme = ttheme(base_style = "lBlueWhite"))
```