---
title: "Tutorial"
output: learnr::tutorial
runtime: shiny_prerendered
---

```{r setup, include=FALSE}
library(learnr)
knitr::opts_chunk$set(echo = FALSE)

set.seed(783623)
library(tidyverse)
experimental_data<-tibble(participant_id = runif(120, min = 10000000, max = 99999999),
  session = sample(factor(x = c("morning", "midday", "evening"), levels = c("morning", "midday", "evening"), ordered = TRUE), size = 120, replace = TRUE, prob = c(.25, .30, .45)),
          Administrator = sample(c("UG", "Graduate"), size = 120, c(.5, .5), replace = TRUE))%>%
  mutate(intervention_finished = case_when(session == "morning" ~ rbinom(n(), 1, .75),
                               session == "midday" ~ rbinom(n(), 1, .80), 
                               session == "evening" ~ rbinom(n(), 1, .5)))%>%
  mutate(memory_task_pre = sample(size = n(), c("Excellent", "Good", "Adequate", "Poor", "Terrible"), replace = TRUE, prob = c(.15, .2, .15, .2, .3)),
    memory_task_post = case_when(intervention_finished == 1 ~ sample(size = n(), c("Excellent", "Good", "Adequate", "Poor", "Terrible"), replace = TRUE, prob = c(.3, .4, .2, .05, .05)),
                            intervention_finished == 0 ~ sample(size = n(), c("Excellent", "Good", "Adequate", "Poor", "Terrible"), replace = TRUE, prob = c(.2, .2, .3, .1, .2))),
    GPA = if_else(session == "evening", rnorm(120, 2, sd =.75), rnorm(120, 2.3, .30)), 
    SWB =(((GPA-mean(GPA))/sd(GPA)*.3+rnorm(120, 0, sd = .91))+2.5)*1.3)
```
# Summary Statistics with Base R and Dplyr

## Q & D Summary Statistics
Use skimr to summarize experimental_data. How does it treat factors?

```{r qd_summary, exercise.lines=15, exercise = TRUE}

```

## Custom Summary Reports
Generate custom reports for numeric variables using the appropriate functions. In the report calculate maximum and minimum values and standard deviations. 


```{r custom, exercise.lines=15, exercise = TRUE}

```


## Measures of Association
Create a correlation table for all numeric variables. In what way is GPA related to subjective well being (i.e., positive, negative, not at all)?

```{r associated, exercise.lines=15, exercise = TRUE}

```


## Grouped Summaries
Create summary report for each session. Report the mean, standard deviation, and frequency. Which session had the highest GPA?

```{r grouped, exercise.lines=15,  exercise = TRUE}

```


## Visualizations of Distributions
Visualize the distributions of GPA across sessions. Which session has the most variability?

```{r visuals, exercise.lines=15, exercise = TRUE}

```